Foundations of ecological and evolutionary change

### 2025-05-31 foundations_of_ecoevo.tex
Draft a rewrite of paragraphs 1 to 3

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\usepackage{lineno}
\linenumbers
\modulolinenumbers[2]
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Darwin's Dream: Unifying ecological and evolutionary change},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Darwin's Dream: Unifying ecological and evolutionary change}
\author{A. Bradley Duthie\(^{1,a,*}\) and Victor J. Luque\(^{2,a}\)}
\date{{[}1{]} Department of Biological and Environmental Sciences,
University of Stirling, Stirling, Scotland {[}2{]} Department of
Philosophy, University of Valencia, Valencia, Spain {[}*{]}
Corresponding author:
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
{[}a{]} Equal contribution}

@@

\setstretch{2}
\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Theoretical unification of phenomena is a powerful tool for scientific
advancement. Such unification has been a major goal in scientific
research throughout history (\protect\hyperlink{ref-Kitcher1993}{Kitcher
1993}), and its value is perhaps most evident in reconciling unconnected
models and revealing new and unexpected empirical predictions. In
theoretical physics, Newton's unification of celestial and terrestrial
motion, and Maxwell's unification of electricity and magnetism, are
among the most important advances in scientific history. Likewise, in
the life sciences, the modern synthesis of natural selection and
Mendelian inheritance revolutionised how mechanisms underlying evolution
and genetics are understood and applied
(\protect\hyperlink{ref-Gayon2019}{Gayon and Huneman 2019}). The analogy
to physics is imperfect because evolution encompasses non-biological
systems with non-genetic inheritance
(\protect\hyperlink{ref-Price1995}{Price 1995};
\protect\hyperlink{ref-Shennan2011}{Shennan 2011};
\protect\hyperlink{ref-MacCallum2012}{MacCallum et al. 2012};
\protect\hyperlink{ref-Lewens2015}{Lewens 2015}), but to illustrate our
point, we restrict our focus to biological evolution. Following the
modern synthesis, patterns in evolutionary biology and genetics could be
interpreted through a common framework.

Selection, mutation, recombination, migration, and drift all have
predictable consequences for both phenotypes and genotypes in a
population. These mechanisms are therefore foundational to understanding
evolution and genetics, and their roles in explaining evolutionary and
genetic phenomena are no more conceptually pluralistic than that of
gravity in explaining celestial and terrestrial motion. Instead,
biologists understand that each of these mechanisms will jointly affect
genetic composition and evolutionary change in a population. For
example, all else being equal, drift will lead to a loss of both
heterozygosity and adaptive phenotypes. This constrains the possible
relations that can exist between genetic composition and evolutionary
change and therefore clarifies the scope of hypotheses that are
biologically coherent (\protect\hyperlink{ref-Caswell1988}{Caswell
1988}). It would obviously be mistaken to model a population in which
drift had inconsistent effects on evolution and genetics, but such
mistakes would surely be common without the foundational theory required
to unify the two subjects. The absence of this theory is almost
unthinkable in modern biology. Evolution and genetics are arguably as
intertwined in biology as electricity and magnetism are in physics.

In contrast, theory underlying ecology and evolution has not been so
tightly intertwined. Until recently, population ecology and evolution in
particular have taken almost parallel paths for developing theory.
Although many revelatory models exist to connect ecological and
evolutionary phenomena (\protect\hyperlink{ref-Lion2018}{Lion 2018};
\protect\hyperlink{ref-Patel2018}{Patel et al. 2018};
\protect\hyperlink{ref-Govaert2019}{Govaert et al. 2019}), the
foundational assumptions and equations of ecology and evolution in these
models remain mostly separated conceptually. But like the mechanisms
driving evolutionary and genetic change, those that drive ecological and
evolutionary change are well-established with predictable joint
consequences. In evolution, natural selection, drift, and gene flow are
all instantiated through a combination of different ecological factors
(\protect\hyperlink{ref-Connor2004}{Connor and Hartl 2004}). Most
notably, change in population density and phenotype are both caused by
individual birth and death (\protect\hyperlink{ref-Turchin2001}{Turchin
2001}). This should allow biologists to place ecological and
evolutionary change on a shared theoretical foundation.

We believe that such a foundation exists for population ecology and
evolutionary biology, and that it can bridge the fundamental equations
of both subjects and serve as a starting point for understanding change
in any population. Turchin (\protect\hyperlink{ref-Turchin2001}{2001})
argues for a simple birth and death model (Box 1) as a fundamental
equation of population ecology, and for exponential growth to
consequently be as fundamental to ecology as Newton's first law of
motion is to physics. Turchin
(\protect\hyperlink{ref-Turchin2001}{2001}) argues that general
principles are needed to establish a logical foundation for population
ecology. In evolutionary biology, the Price equation (Box 2) is almost
universally regarded as fundamental
(\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Rice2004}{Rice 2004};
\protect\hyperlink{ref-Gardner2008}{Gardner 2008};
\protect\hyperlink{ref-Frank2015a}{Frank 2017};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Luque2021}{Luque and Baravalle 2021}). It
exhaustively and exactly describes evolutionary change for any closed
evolving population (\protect\hyperlink{ref-Price1970}{Price 1970};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Lehtonen2020}{Lehtonen et al. 2020}), and it
serves as the starting point for all other fundamental equations in
evolution (\protect\hyperlink{ref-Queller2017}{Queller 2017}). We
believe that this logical foundation can be made sufficiently broad to
encompass both population and evolutionary change.

We propose an equation from which the fundamental ecological and
evolutionary equations can be derived. Derivation follows by adding
assumptions that are specific to population ecology or evolution in the
same way that key equations of population genetics (e.g., average
excess) or quantitative genetics (e.g., Breeder's equation) can be
derived from the Price equation
(\protect\hyperlink{ref-Queller2017}{Queller 2017}). We hope that our
equation can provide a starting point for building a shared logical
framework between ecology and evolution.

@@

\begin{quote}
\textbf{Box 1:} The number of individuals in any closed population
(\(N\)) at any given time (\(t + 1\)) is determined by the existing
number of individuals (\(N_{t}\)), plus the number of births
(\(Births\)) minus the number of deaths (\(Deaths\)),
\[N_{t+1} = N_{t} + Births - Deaths.\] This equation is necessarily true
for any closed population. Despite its simplicity, it is a general
equation for defining population change and a starting point for
understanding population ecology. Turchin
(\protect\hyperlink{ref-Turchin2001}{2001}) notes that a consequence of
this fundamental equation is the tendency for populations to grow
exponentially (technically geometrically in the above case where time is
discrete). This inherent underlying tendency towards exponential growth
persists even as the complexities of real populations, such as
structure, stochasticity, or density-dependent effects are added to
population models (\protect\hyperlink{ref-Turchin2001}{Turchin 2001}).
Given the assumption that all individuals in the population are
identical, a per capita rate of birth, \(Births = bN_{t}\) and death,
\(Deaths = dN_{t}\) can be defined. Rearranging and defining
\(\lambda = 1 + b - d\) gives, \(N_{t+1} = N_{t}\lambda\). Here
\(\lambda\) is the finite rate of increase
(\protect\hyperlink{ref-Gotelli2001}{Gotelli 2001}), and note that
because \(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in
size of any closed population equals its existing size times its finite
rate of increase.
\end{quote}

@@

\begin{quote}
\textbf{Box 2:} The Price equation is an abstract formula to represent
evolutionary change. Formulated originally in the early 1970s by George
Price (\protect\hyperlink{ref-Price1970}{Price 1970},
\protect\hyperlink{ref-Price1972}{1972}), it postulates some basic
properties that all evolutionary systems must satisfy: change over time,
ancestor and descendant relations, and a character or phenotype
(\protect\hyperlink{ref-Rice2004}{Rice 2004}). Using simple algebraic
language, the Price equation represents evolutionary change with the
predominant notation,
\[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
In the above equation, \(\Delta\bar{z}\) is the change in the average
character value \(z\) over a time step of arbitrary length, \(w\) is an
individual's absolute fitness, and \(\bar{w}\) average population
fitness. On the right-hand side of the equation, the first term is the
covariance between a character value \(z\) and fitness \(w\), which
reflects \(\bar{z}\) change attributable to differential survival and
reproduction. The second term is the expected value of \(\Delta z\),
which reflects the extent to which offspring deviate from parents in
\(z\) (\protect\hyperlink{ref-Rice2004}{Rice 2004};
\protect\hyperlink{ref-Okasha2006}{Okasha 2006};
\protect\hyperlink{ref-Frank2012}{Frank 2012}). A more specific version
of the covariance term was already known within the quantitative and
population genetics tradition
(\protect\hyperlink{ref-Robertson1966}{Robertson 1966}), usually
representing the action of natural selection. The Price equation adds an
expectation term and abstracts away from any specific mechanisms of
replication or reproduction, or mechanisms of inheritance. Its
definitional nature and lack of substantive biological assumptions has
been portrayed both as a strength (\protect\hyperlink{ref-Rice2004}{Rice
2004}; \protect\hyperlink{ref-Frank2012}{Frank 2012};
\protect\hyperlink{ref-Luque2016}{Luque 2017};
\protect\hyperlink{ref-Baravalle2022}{Baravalle and Luque 2022}), and
its greatest weakness. The abstract nature of the Price equation places
it at the top of the hierarchy of fundamental theorems of evolution from
which the rest (Robertson's theorem, Fischer's fundamental theorem,
breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
easily derived (\protect\hyperlink{ref-Lehtonen2016}{Lehtonen 2016},
\protect\hyperlink{ref-Lehtonen2018}{2018};
\protect\hyperlink{ref-Queller2017}{Queller 2017}). This abstractness is
also key to developing a more general view of evolution
(\protect\hyperlink{ref-Rice2020}{Rice 2020};
\protect\hyperlink{ref-Luque2021}{Luque and Baravalle 2021};
\protect\hyperlink{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast,
some researchers consider the Price equation just a triviality (even
tautological), and useless without further modelling assumptions
(\protect\hyperlink{ref-VanVeelen2005}{van Veelen 2005};
\protect\hyperlink{ref-VanVeelen2012}{van Veelen et al. 2012}). The
debate remains open (\protect\hyperlink{ref-VanVeelen2020}{van Veelen
2020}; \protect\hyperlink{ref-Baravalle2024}{Baravalle et al. 2024}).
\end{quote}

@@

\hypertarget{a-foundation-for-biological-evolution-and-population-ecology}{%
\section{A foundation for biological evolution and population
ecology}\label{a-foundation-for-biological-evolution-and-population-ecology}}

To fully unify biological evolution and population ecology, we must
reconcile the general equation for population change (Box 1) with the
Price equation (Box 2). Previous theory has produced mathematical
synthesis between ecological and evolutionary equations
(\protect\hyperlink{ref-Page2002}{Page and Nowak 2002};
\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), but not
established a unified conceptual framework for joint ecological and
evolutionary change. For example, Page and Nowak
(\protect\hyperlink{ref-Page2002}{2002}) demonstrate a mathematical
equivalence between the Price equation and a general equation for
community dynamics. In both equations, the relative frequencies of
different ecological (species) or evolutionary (phenotype) variants map
to their respective ecological (population growth) or evolutionary
(fitness) change in a way that is mathematically equivalent. Collins and
Gardner (\protect\hyperlink{ref-Collins2009}{2009}) use the Price
equation to build a general quantitative framework that partitions
phenotypic change into ecological, evolutionary, and physiological
components. The average change in a community-wide trait (e.g., carbon
uptake) is exactly and generally partitioned into change attributable to
changes in species composition, expected evolution of lineages within
species, and expected change in lineage physiology across all species.
Both Page and Nowak (\protect\hyperlink{ref-Page2002}{2002}) and Collins
and Gardner (\protect\hyperlink{ref-Collins2009}{2009}) use relative
species abundances, meaning that species frequencies must sum to one.
This is useful for demonstrating mathematical equivalence
(\protect\hyperlink{ref-Page2002}{Page and Nowak 2002}) or expected
change in a community-wide trait
(\protect\hyperlink{ref-Collins2009}{Collins and Gardner 2009}), but a
different approach is needed to recover absolute change in population
size and thereby recover a joint foundation for ecology and evolution.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{fancyhdr}
\usepackage{lineno}
\linenumbers
\modulolinenumbers[2]
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Foundations of ecological and evolutionary change},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Foundations of ecological and evolutionary change}
\author{A. Bradley Duthie\(^{1,a,*}\) and Victor J. Luque\(^{2,a}\)}
\date{{[}1{]} Department of Biological and Environmental Sciences,
University of Stirling, Stirling, Scotland {[}2{]} Department of
Philosophy, University of Valencia, Valencia, Spain {[}*{]}
Corresponding author:
\href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}
{[}a{]} Equal contribution}

@@

\setstretch{2}
\section{Abstract}\label{abstract}

@@

\section{Introduction}\label{introduction}

Theoretical unification of phenomena is a powerful tool for scientific
advancement. Such unification has been a major goal in scientific
research throughout history (\citeproc{ref-Kitcher1993}{Kitcher 1993}),
and its value is perhaps most evident in reconciling unconnected models
and revealing new and unexpected empirical predictions. In theoretical
physics, Newton's unification of celestial and terrestrial motion, and
Maxwell's unification of electricity and magnetism, are among the most
important advances in scientific history. Likewise, in the life
sciences, the modern synthesis of natural selection and Mendelian
inheritance revolutionised how mechanisms underlying evolution and
genetics are understood and applied (\citeproc{ref-Gayon2019}{Gayon and
Huneman 2019}). The analogy to physics is imperfect because evolution
encompasses non-biological systems with non-genetic inheritance
(\citeproc{ref-Price1995}{Price 1995};
\citeproc{ref-Shennan2011}{Shennan 2011};
\citeproc{ref-MacCallum2012}{MacCallum et al. 2012};
\citeproc{ref-Lewens2015}{Lewens 2015}), but to illustrate our point, we
restrict our focus to biological evolution. Following the modern
synthesis, patterns in evolutionary biology and genetics could be
interpreted through a common framework.

Selection, mutation, recombination, migration, and drift all have
predictable consequences for both phenotypes and genotypes in a
population. These mechanisms are therefore foundational to understanding
evolution and genetics, and their roles in explaining evolutionary and
genetic phenomena are no more conceptually pluralistic than that of
gravity in explaining celestial and terrestrial motion. Instead,
biologists understand that each of these mechanisms will jointly affect
genetic composition and evolutionary change in a population. For
example, all else being equal, drift will lead to a loss of both
heterozygosity and adaptive phenotypes. This constrains the possible
relations that can exist between genetic composition and evolutionary
change and therefore clarifies the scope of hypotheses that are
biologically coherent (\citeproc{ref-Caswell1988}{Caswell 1988}). It
would obviously be mistaken to model a population in which drift had
inconsistent effects on evolution and genetics, but such mistakes would
surely be common without the foundational theory required to unify the
two subjects. The absence of this theory is almost unthinkable in modern
biology. Evolution and genetics are arguably as intertwined in biology
as electricity and magnetism are in physics.

In contrast, theory underlying ecology and evolution has not been so
tightly intertwined. Until recently, population ecology and evolution in
particular have taken almost parallel paths for developing theory.
Although many revelatory models exist to connect ecological and
evolutionary phenomena (\citeproc{ref-Lion2018}{Lion 2018};
\citeproc{ref-Patel2018}{Patel et al. 2018};
\citeproc{ref-Govaert2019}{Govaert et al. 2019}), the foundational
assumptions and equations of ecology and evolution in these models
remain mostly separated conceptually. But like the mechanisms driving
evolutionary and genetic change, those that drive ecological and
evolutionary change are well-established with predictable joint
consequences. In evolution, natural selection, drift, and gene flow are
all instantiated through a combination of different ecological factors
(\citeproc{ref-Connor2004}{Connor and Hartl 2004}). Most notably, change
in population density and phenotype are both caused by individual birth
and death (\citeproc{ref-Turchin2001}{Turchin 2001}). This should allow
biologists to place ecological and evolutionary change on a shared
theoretical foundation.

We believe that such a foundation exists for population ecology and
evolutionary biology, and that it can bridge the fundamental equations
of both subjects and serve as a starting point for understanding change
in any population. Turchin (\citeproc{ref-Turchin2001}{2001}) argues for
a simple birth and death model (Box 1) as a fundamental equation of
population ecology, and for exponential growth to consequently be as
fundamental to ecology as Newton's first law of motion is to physics.
Turchin (\citeproc{ref-Turchin2001}{2001}) argues that general
principles are needed to establish a logical foundation for population
ecology. In evolutionary biology, the Price equation (Box 2) is almost
universally regarded as fundamental (\citeproc{ref-Price1970}{Price
1970}; \citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Gardner2008}{Gardner 2008};
\citeproc{ref-Frank2015a}{Frank 2017}; \citeproc{ref-Luque2016}{Luque
2017}; \citeproc{ref-Luque2021}{Luque and Baravalle 2021}). It
exhaustively and exactly describes evolutionary change for any closed
evolving population (\citeproc{ref-Price1970}{Price 1970};
\citeproc{ref-Luque2016}{Luque 2017};
\citeproc{ref-Lehtonen2020}{Lehtonen et al. 2020}), and it serves as the
starting point for all other fundamental equations in evolution
(\citeproc{ref-Queller2017}{Queller 2017}). We believe that this logical
foundation can be made sufficiently broad to encompass both population
and evolutionary change.

We propose an equation from which the fundamental ecological and
evolutionary equations can be derived. Derivation follows by adding
assumptions that are specific to population ecology or evolution in the
same way that key equations of population genetics (e.g., average
excess) or quantitative genetics (e.g., Breeder's equation) can be
derived from the Price equation (\citeproc{ref-Queller2017}{Queller
2017}). We hope that our equation can provide a starting point for
building a shared logical framework between ecology and evolution.

@@

\begin{quote}
\textbf{Box 1:} The number of individuals in any closed population
(\(N\)) at any given time (\(t + 1\)) is determined by the existing
number of individuals (\(N_{t}\)), plus the number of births
(\(Births\)) minus the number of deaths (\(Deaths\)),
\[N_{t+1} = N_{t} + Births - Deaths.\] This equation is necessarily true
for any closed population. Despite its simplicity, it is a general
equation for defining population change and a starting point for
understanding population ecology. Turchin
(\citeproc{ref-Turchin2001}{2001}) notes that a consequence of this
fundamental equation is the tendency for populations to grow
exponentially (technically geometrically in the above case where time is
discrete). This inherent underlying tendency towards exponential growth
persists even as the complexities of real populations, such as
structure, stochasticity, or density-dependent effects are added to
population models (\citeproc{ref-Turchin2001}{Turchin 2001}). Given the
assumption that all individuals in the population are identical, a per
capita rate of birth, \(Births = bN_{t}\) and death, \(Deaths = dN_{t}\)
can be defined. Rearranging and defining \(\lambda = 1 + b - d\) gives,
\(N_{t+1} = N_{t}\lambda\). Here \(\lambda\) is the finite rate of
increase (\citeproc{ref-Gotelli2001}{Gotelli 2001}), and note that
because \(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in
size of any closed population equals its existing size times its finite
rate of increase.
\end{quote}

@@

\begin{quote}
\textbf{Box 2:} The Price equation is an abstract formula to represent
evolutionary change. Formulated originally in the early 1970s by George
Price (\citeproc{ref-Price1970}{Price 1970},
\citeproc{ref-Price1972}{1972}), it postulates some basic properties
that all evolutionary systems must satisfy: change over time, ancestor
and descendant relations, and a character or phenotype
(\citeproc{ref-Rice2004}{Rice 2004}). Using simple algebraic language,
the Price equation represents evolutionary change with the predominant
notation,
\[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
In the above equation, \(\Delta\bar{z}\) is the change in the average
character value \(z\) over a time step of arbitrary length, \(w\) is an
individual's absolute fitness, and \(\bar{w}\) average population
fitness. On the right-hand side of the equation, the first term is the
covariance between a character value \(z\) and fitness \(w\), which
reflects \(\bar{z}\) change attributable to differential survival and
reproduction. The second term is the expected value of \(\Delta z\),
which reflects the extent to which offspring deviate from parents in
\(z\) (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Okasha2006}{Okasha 2006}; \citeproc{ref-Frank2012}{Frank
2012}). A more specific version of the covariance term was already known
within the quantitative and population genetics tradition
(\citeproc{ref-Robertson1966}{Robertson 1966}), usually representing the
action of natural selection. The Price equation adds an expectation term
and abstracts away from any specific mechanisms of replication or
reproduction, or mechanisms of inheritance. Its definitional nature and
lack of substantive biological assumptions has been portrayed both as a
strength (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Frank2012}{Frank 2012}; \citeproc{ref-Luque2016}{Luque
2017}; \citeproc{ref-Baravalle2022}{Baravalle and Luque 2022}), and its
greatest weakness. The abstract nature of the Price equation places it
at the top of the hierarchy of fundamental theorems of evolution from
which the rest (Robertson's theorem, Fisher's fundamental theorem,
breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
easily derived (\citeproc{ref-Lehtonen2016}{Lehtonen 2016},
\citeproc{ref-Lehtonen2018}{2018}; \citeproc{ref-Queller2017}{Queller
2017}). This abstractness is also key to developing a more general view
of evolution (\citeproc{ref-Rice2020}{Rice 2020};
\citeproc{ref-Luque2021}{Luque and Baravalle 2021};
\citeproc{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast, some
researchers consider the Price equation just a triviality (even
tautological), and useless without further modelling assumptions
(\citeproc{ref-VanVeelen2005}{van Veelen 2005};
\citeproc{ref-VanVeelen2012}{van Veelen et al. 2012}). The debate
remains open (\citeproc{ref-VanVeelen2020}{van Veelen 2020};
\citeproc{ref-Baravalle2024}{Baravalle et al. 2024}).
\end{quote}

@@

\section{A foundation for biological evolution and population
ecology}\label{a-foundation-for-biological-evolution-and-population-ecology}

To fully unify biological evolution and population ecology, we must
reconcile the general equation for population change (Box 1) with the
Price equation (Box 2). Previous theory has produced mathematical
synthesis between ecological and evolutionary equations
(\citeproc{ref-Page2002}{Page and Nowak 2002};
\citeproc{ref-Collins2009}{Collins and Gardner 2009}), but not
established a unified conceptual framework for joint ecological and
evolutionary change. For example, Page and Nowak
(\citeproc{ref-Page2002}{2002}) demonstrate a mathematical equivalence
between the Price equation and a general equation for community
dynamics. In both equations, the relative frequencies of different
ecological (species) or evolutionary (phenotype) variants map to their
respective ecological (population growth) or evolutionary (fitness)
change in a way that is mathematically equivalent. Collins and Gardner
(\citeproc{ref-Collins2009}{2009}) use the Price equation to build a
general quantitative framework that partitions phenotypic change into
ecological, evolutionary, and physiological components. The average
change in a community-wide trait (e.g., carbon uptake) is exactly and
generally partitioned into change attributable to changes in species
composition, expected evolution of lineages within species, and expected
change in lineage physiology across all species. Both Page and Nowak
(\citeproc{ref-Page2002}{2002}) and Collins and Gardner
(\citeproc{ref-Collins2009}{2009}) use relative species abundances,
meaning that species frequencies must sum to one. This is useful for
demonstrating mathematical equivalence (\citeproc{ref-Page2002}{Page and
Nowak 2002}) or expected change in a community-wide trait
(\citeproc{ref-Collins2009}{Collins and Gardner 2009}), but a different
approach is needed to recover absolute change in population size and
thereby recover a joint foundation for ecology and evolution.

The Price equation is critical for partitioning different components of
biological change (\citeproc{ref-Price1970}{Price 1970};
\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Gardner2008}{Gardner
2008}; \citeproc{ref-Luque2016}{Luque 2017};
\citeproc{ref-Queller2017}{Queller 2017};
\citeproc{ref-Lehtonen2018}{Lehtonen 2018}). It has also been highly
useful for integrating evolutionary theory across disciplines
(\citeproc{ref-Fox2006}{Fox 2006};
\citeproc{ref-Brantingham2007}{Brantingham 2007};
\citeproc{ref-MacCallum2012}{MacCallum et al. 2012};
\citeproc{ref-Frank2015}{Frank 2015}; \citeproc{ref-Godsoe2021}{Godsoe
et al. 2021}; \citeproc{ref-Ulrich2024}{Ulrich et al. 2024}). These
properties would seem to make it an intuitive starting point for a
logical foundation of ecology and evolution, perhaps through some kind
of mathematical equivalence (\citeproc{ref-Page2002}{Page and Nowak
2002}) or addition of terms (\citeproc{ref-Collins2009}{Collins and
Gardner 2009}), or through the use of its recursive structure
(\citeproc{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
\citeproc{ref-Frank2012}{Frank 2012}). But despite its flexibility, the
Price equation still relies on relative frequencies, which must by
definition sum to one (\citeproc{ref-Frank2015}{Frank 2015}). This is
because the Price equation describes the average change in a population;
the frequency of entities is scaled thereby conserving total probability
(\citeproc{ref-Frank2015}{Frank 2015}, \citeproc{ref-Frank2016}{2016}).
But to recover the fundamental principle of exponential population
growth (\citeproc{ref-Turchin2001}{Turchin 2001}), this scaling must be
avoided in a fundamental equation of ecology and evolution.
